\todo{INGRID: Creating Languages in MPS by Importing ANTLR Grammars}

### 2016-09-23 paper.tex
Split paper into chapter files

% TODO pridat nekoho dalsiho z JetBrains
% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
\authorinfo{V\'aclav Pech}
	{JetBrains}

@@

\section{Introduction}

\JB{Odstavec mirne upraven} The prevailing approach to define valid syntaxes for programming languages is through grammars written typically in notations based on (or similar to) EBNF [reference]. Many popular tools exist for automated generation of parsers from the grammar definitions, such as ANTLR and Bison/Flex. The source code of programs in such languages is written in a text editors, usually embedded into an IDE.

\JB{Odstavec mirne upraven} Projectional (or structural) editing offers an alternative approach. While the idea emerged as early as in 1970s, it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.  A recent revival of projectional editing has been observed in the area of language workbenches - IDE-like tools that enable the developers to manipulate the actual language definition. The Meta-programming System (MPS) [reference] by JetBrains is an open-source language workbench that focuses of Domain Specific Languages and leverages the concept of projectional editor. MPS allows developers to design custom languages and then use use them to write code and compile it into executables. MPS provides the whole IDE infrastructure for such languages.

\JB{Odstavec mirne upraven} The key idea behind projectional editors is that a developer directly manipulates the AST representation of the source code instead of plain text. To avoid the fate of the early projectional editors MPS has put extra effort into making text manipulation feel like editing real text. MPS models, although being persisted as XML or binary files, can be versioned using the industry standard source-control systems (Git, Subversion). On-line model browsers have been incorporated into some of the popular public source code repositories (BitBucket) [reference] .

\JB{Odstavec mirne upraven} The absence of parsers brings about numerous benefits, which are in particular useful for DSLs, which are frequently used by domain experts, who may not be professional programmers.
\begin{itemize}
	\item Since a projectional editors controls the interaction between the user and the code, the developers practically cannot enter syntactically invalid code.
	\item Languages can be defined in a modular way and multiple languages can be easily combined together or extend one another, while parsers seriously limit language modularity for traditional parsed languages.
	\item The languages may offer contextual or non-parseable notations, such as tables, diagrams or positional syntax.
	\item Since the projection is detached from the physical representation of code (AST), authors can define multiple notations and allow the developers to switch between different representations of the code.
\end{itemize}
A motivating illustration of the abilities of MPS is the mbeddr project [citation], which extends C with numerous domain-specific constructs and data types.
We provide more details about MPS in Section 2.1.

\JB{Navrhuji smazat nasledujici odstavec}
\todo{From thesis.
	One of these tools is the JetBrains MPS~\cite{ref:MPS}, which approaches languages from a~different point of view.
	MPS doesn't use the textual representation of code as usual, but rather works with the actual abstract syntax tree that holds the code's structure.
	This has some positive implications as it gives us a~very powerful tool rich on features but also introduces some new problems we haven't encountered before.
	One of them being the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them.
	MPS also becomes a~very powerful tool once the language has already been created and you start using it inside MPS.
	There is a~huge variety of new languages constructed for MPS that usually solve domain specific problems.
}

\todo{Pouzivat jen terminy "AST nodes" a "language elements" (to spis). Nepouzivat pojem "concepts" abychom ctenare nematli.}

\JB{Odstavec upraven} Projectional editing introduces some new problems we haven't encountered before. One of them is the need for a~special definition of languages created inside MPS so that the IDE can understand them and work with them. In order to use a language inside MPS, it has to be defined through the MPS infrastructure. In particular, one has to create the abstract syntax called Structure (the types of allowed AST nodes), concrete syntax called Editor (projection of the AST on the screen and its interactions with the user), and text generation (to enable creation of plain text files used as input for compilers). We provide more details in Section X (Background about MPS).

\JB{Odstavec pridan pro vysvetleni motivace prace} MPS is typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages. These DSLs are then transformed on the AST level by a series of model-to-model transformations to an AST model representing the desired semantics in a General-purpose language (GPL), such as C or Java. This model is then converted to a textual representation of that language and compiled by the standard means of the target platform. This implies that the target GPL must also be defined by the MPS means, to allow the DSLs to have their models transformed into a model in the GPL.

\JB{Odstavec doplnen} However, it requires substantial effort to create a definition of a GPL in MPS. Therefore, very few mainstream GPLs are now fully supported by MPS. Only Java and C have been implemented to date, which limits the DSLs authors to only target these supported target GPLs. We believe that an automated process of importing grammars of GPLs into MPS would encourage more GPLs to be migrated to MPS and thus giving the DSL authors more options.

\todo{From thesis.
	However, building these languages is a~complicated and time-consuming process and a~lot of effort must be put into them before they are ready for use.

	These examples show that recreating a~full language inside MPS is not an easy task and a~lot of time must be spent on implementing all aspects of the language.
	A big part of this effort is, however, quite straightforward and can be possibly automated, as we show in this paper. Automation speeds up the process of adoption of new languages to a great degree.
}

\JB{Odstavec upraven} Since a big part of the language definition process in MPS is quite straightforward and can possibly be automated, we will look into the possibility of automatic import of already existing general purpose languages into MPS using a grammar description of their syntax.

We present INGRID --- a method and a tool for semi-automated construction of MPS definition of programming languages based on their ANTLR grammars.
We also discuss challenges that we encountered during this process and our solution to them. Most of these challenges are general, i.e. not MPS specific, as they apply to the relation between grammars defined in EBNF-like style (rules and tokens, everything defined in plain text) and structured (object-oriented) approach used in MPS (and other language workbenches).

\todo{Hodne zduraznit automation, tedy ze chceme maximalizovat automatizaci.}
The main goal of our/the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible.

\todo{From thesis.
	Automatic import of already existing languages into MPS from the grammar description of their syntax.
}
\todo{Mozna prodat INGRID jako automatickou konverzi gramatik do podoby MPS languages, tedy jako nastroj ktery umoznuje portaci jazyku do prostredi MPS a jejich nasledne pouziti v MPS (treba kombinace s dalsimi jazyky).}

Our primary motivation is to enable better support for mainstream and/or general purpose programming languages (XML, C, C++, JavaScript, Python) in MPS, especially to enable writing programs in these languages using the projectional editor and to combine them with domain-specific languages natively created in MPS.
\todo{Zduraznit take pouziti importovanych jazyku pro cile generovani z DSL --- to plati zejmena u mainstream general-purpose languages jako treba JavaScript.}

\todo{Briefly mention the biggest challenges.}
The are related to the way grammar encoding/description of languages are usually written.
They do not hold any information about the code layout, and contain many rules that do not directly correspond to AST nodes (programming language constructs).
We discuss them in more detail later (Section X).

\todo{From thesis.
	INGRID allows developers to leverage all the features that MPS has to offer together with general purpose languages such as C++, JavaScript or Python.
	The biggest benefit is the possibility to code in these mainstream general purpose languages using the projectional editor.
}

However, we do not target just the very complex general purpose languages such as JavaScript.
INGRID can be used also for simple languages.
We believe that, for simple languages, it is less time consuming to create the grammar by hand (e.g., using the ANLTR notation and toolset) and then create the language definition in MPS by using the INGRID tool, than to create it manually using the MPS GUI --- this is another benefit of INGRID.

Structure of the paper.
In the beginning, we will describe the MPS editor and related/similar projects.
Then, we describe our approach to creating language definition in MPS from their grammars, and discuss the major challenges together with our solution.
We will also show some example languages imported with INGRID and discuss our experience.

\section{Background and Related Work}

We will briefly introduce fundamental basics of the MPS environment.
We will analyze existing related projects that are trying to accomplish similar goals, and then we will weigh advantages and disadvantages of these approaches.

\subsection{JetBrains MPS}

\JB{Odstavec upraven} JetBrains MPS is a complete language workbench - a development environment that allows developers to define their own languages and then use them to develop code. The code can then be transformed into a target language, typically a GPL, such as Java or C.

\JB{Navrhuji odstranit} We need to provide a good understanding of how the MPS~\cite{ref:MPS} environment works.
Focus on basic/fundamental and relevant features. In this section, we will explain fundamental basics of MPS as it is crucial in the understanding of our contribution and associated challenges.

\JB{odstavec upraven} MPS differs from typical IDEs in one important aspect - the projectional editor. The developer does not work with the textual representation of the source code directly, but rather with its AST (abstract syntax tree) that is the model of the code. When users are creating programs using the MPS projectional editor, they are basically assembling the tree (AST) out of predefined building blocks of the imported languages. The definition of the language dictates, which where in the AST can certain elements be placed or how they can be nested inside other elements. In traditional IDEs it is the parser that constructs an AST using the language's elements.

\JB{Odstavec upraven} The building blocks of MPS models are called nodes. Code of any program in MPS is built from nodes and these nodes represent instances of elements defined in one of the languages that the program declares to be using. These language elements are called concepts. In brief, nodes in code are instances of concepts defined in the language.

\JB{Vyhody/nevyhody diskutujeme v kapitole 1. Navrhuji nasledujici odstavec zrusit nebo text z kapitoly 1. presunout sem}
Keeping the model of the code in the AST form has several advantages.
One of them is that MPS then only allows composing the program strictly following the syntax of the language, which results in an inability to actually write syntactically invalid code.
The code is generated out of the always-valid AST beyond user's reach.

\JB{Odstavec upraven} One of key advantages of projectional editor stems from separation of abstract and concrete syntax. While the AST provides an exhaustive representation of the code, the way it is displayed on the screen and the way the user interacts with it are unconstrained. The editor can take any visual form and shape. The language author can define multiple visualizations and let the developer to choose one that fits best the task at hands - editing, debugging, reviewing, resolving merge conflicts, etc.

For a better understanding, we will give a~small example. Let's assume we have a language with typical if/else structure similar to the one in Java:

\begin{center}
	\begin{minipage}{.38\textwidth}
		\begin{alltt}
			if (\textit{condition})
			    \textit{then statement block}
			else
			    \textit{else statement block}
		\end{alltt}
	\end{minipage}
\end{center}

Its abstract syntax tree representation will look something like shown on Figure~\ref{fig:if_ast}.

\begin{figure}[ht]
  \centering
  \includegraphics[scale=0.6]{./images/if_statement_ast.png}
  \caption{"If statement" abstract syntax tree}
  \label{fig:if_ast}
\end{figure}

The underlying child nodes have their own children and so on. The type of nodes can be restricted by the language structure definition. For example, we can restrict the condition in the if/else statement to be an expression, statement block to be a list of statements, etc.

\JB{odstavec upraven} Alternative visual representations are possible and they are not bound to be just the textual representation at all. For example, blocks for both of the branches can be aligned next to each other and displayed with differend background color, like shown in Figure~\ref{fig:if_editor}.

\begin{figure}[ht]
	\centering
	\hspace{-4mm}
	\includegraphics[scale=0.75]{./images/if_statement_editor.png}
	\caption{"If statement" projectional editor example}
	\label{fig:if_editor}
\end{figure}

\JB{Odstavec mirne upraven}The definition of a language element (the concept) consists of several aspects and each aspect codifies a different part of the nodes' behavior. The essential aspects are the following: structure (abstract syntax), editor (concrete syntax - how the code is projected and edited) and textgen (how to transform the AST into textual source code). If, instead of generating text directly, the language is supposed to be generated into another language that is available in MPS, the generator aspect must be used to specify the model-transformation rules. The structure, editor and textgen aspects are most relevant for the contribution of this paper, so we describe them here (and neglect/ignore the other aspects such as type system and data-flow).

\todo{Highlight the object-oriented hierarchical nature of language definitions in MPS. The environment restricts and guides the developer so that he makes less mistakes.}

Some concepts and design patterns known from object oriented programming can be used in the language definition --- inheritance, interfaces, and so on.
The AST node can be inherited from a parent (abstract) node and it can implement a special interface node.
Most importantly, nodes have child nodes and that is how the AST itself is created.

\todo{Zduraznit ze AST node (object) ma nekolik aspectu: structure, editor, textgen. Definice MPS language je organizovana takto --- pres AST nodes, ktere maji nekolik aspectu (ruzne pohledy).}

\paragraph{Structure.}
The most important aspect of any MPS language is structure. It must be created first for each AST node.
It specifies core attributes such as name of the AST node, inheritance relationships, possible child nodes (types and cardinality), implemented interfaces, references to other nodes, other properties (fields of any type that can hold values), etc.
In Figure~\ref{fig:if_statement_structure}, you can see the structure aspect definition for the if statement.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.55]{./images/if_statement_structure.png}
	\caption{"If statement" structure aspect definition}
	\label{fig:if_statement_structure}
\end{figure}

\paragraph{Editor.}
Editor aspect is where the user defines what the projectional editor representation of a code fragment (an AST node) looks like.
The editor aspect of an AST node defines its visual representation (of the code) in the projectional editor.
The user usually incorporates all children, references, and properties inside the representation so that future user of the language can insert their value into these placeholders.
All elements can be styled using a~language similar to CSS.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.5]{./images/if_statement_editor_definition.png}
	\caption{"If statement" editor aspect definition}
	\label{fig:if_editor_definition}
\end{figure}

\todo{From thesis.

The editor is a module dedicated/responsible for the definition of appearance of an AST node (concept).
JetBrains have developed a cellular system that allows placing node's (concept's) properties and children into different cells.
These cells then can be styled to user's liking.
There are many different types of cells, each behaving a little bit different towards its contents.
For example, MPS supports various horizontal and vertical lists, etc.
Extra cells can be added on top --- cells which just specify layout adjustments such as indentation and text color.
In Figure~\ref{fig:if_editor_definition}, you can see what a real editor definition might look like for the if statement of the Java language (as defined in MPS).
}

\paragraph{BaseLanguage}
\todo{Introduce the base language here.}
A program written in an MPS language cannot be executed directly.
MPS uses an approach based on translation into a program in the so-called BaseLanguage.
A plain-text representation of a BaseLanguage program is then passed to a system compiler and the result is executed.
The module responsible for generating the program in the BaseLanguage is called "textgen" (we describe it briefly in the next paragraph).

BaseLanguage is also an MPS language --- it is the official port of Java for MPS (created by JetBrains, authors of MPS).
The default base language of MPS is Java, but others are supported too.

Since BaseLanguage is an MPS language, BaseLanguage code is not a (plain) text source code, but again, an AST built out of nodes belonging to the BaseLanguage language.
This means that generating code inside MPS consists of generating some AST out of BaseLanguage nodes.

Besides representing the default (most common) target of text generation, BaseLanguage is used also for all additional programming inside MPS.

\paragraph{TextGen.}
The TextGen aspect defines how a certain AST node will be translated into a text source code (in the base language).
This will allow us to generate source code in text form out of an AST model and ultimately to generate a text program out of the code user writes in the (imported) MPS language.
The TextGen aspect is basically just one method/procedure definition for each AST node in the language. The method produces string representation of the given AST node and its children.
MPS then calls this method for the root AST node of the program.
TextGen aspect has to be defined (especially its more complex functionality) using the BaseLanguage.
Again, we included an example of the if statement (Figure~\ref{fig:if_statement_textgen}).

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.6]{./images/if_statement_textgen.png}
	\caption{"If statement" TextGen aspect definition}
	\label{fig:if_statement_textgen}
\end{figure}

\todo{From thesis.

After the program's AST is created using the projectional editor, there can be rules defined on how to generate code in plain text out of that tree.
This module is called "text gen".
These defined generators can target any language and platform.

\paragraph{Other aspects.}
MPS supports also other aspects, such as type checking and data-flow analysis, but they are not important for our contribution.
Moreover, information needed for their automated generation is not available/contained inside a grammar file.

}

\subsection{Similar Projects}

Currently, there exists an almost full port of the Java language called BaseLanguage~\cite{ref:BaseLanguage} extended with some MPS specific features.
It was imported manually by JetBrains and it is still undergoing changes as Java itself is evolving.

Within the mbeddr project~\cite{ref:mbeddr}, the C language is also manually tailored for MPS.

Now/here we describe three similar projects (with similar goals) that we are aware of.
We focus on (consider) these three projects: PE4MPS, ANTLR{\_}MPS, and mps-metabnf.

\paragraph{PE4MPS.}
PE4MPS~\cite{ref:PE4MPS} is a project by \todo{Doplnit autora} that is trying to solve the same problem as we do.
It solves the lack of information about code layout in grammars by creating a new grammar notation called PE Grammars~\cite{ref:PE} (the PE abbreviation comes from projectional editing).
It has tw components: PE parser and PE4MPS plugin for MPS.
Their approach is to mimic an existing grammar notation called ANTLRv4~\cite{ANTLR4} and enrich its syntax with custom (their own) constructs.
These extensions tell (guide, hint) the parser what the AST node layout should look like.
Supported extensions (already implemented) include horizontal and vertical lists, and some indentation rules --- however, even these few features make the already not-so-simple syntax of ANTLR v4 much more complex.
The parser then uses this information when generating the projectional editor for am AST node.
Author of this approach/project described the PE syntax using the ANTLRv4 notation~\cite{ANTLR4reference} and then used the ANTLR toolset to automatically generate an ANTLRv4 parser for PE grammar files.
The PE parser reads any PE file and stores the language structure found inside to a custom representation of Java objects.

The PE4MPS plugin/project/tool is built on top of the PE parser.
This plugin uses the PE parser to build the PE file representation (the aforementioned tree-like structure of Java objects) and then creates AST nodes and their aspects inside MPS.
The extended syntax brought by PE describes the layout of each element, e.g. it tells the plugin that one set of child nodes should be displayed horizontally, another set should be vertical with each child on a separate line and with some indentation, and so on.

A limitation of the PE4MPS approach is that it, from what we understand, does not implement full ANTLR syntax.
This means that every grammar might need a non-trivial adjustment before its usage.
One of our goals is to enable import of as many languages out-of-the-box as possible and adopting the full specification.

\paragraph{ANTLR{\_}MPS.}
ANTLR{\_}MPS~\cite{ANTLR2MPS} is another project that is dealing with a similar problem.
The author of this project is Fabien Campagne
The ANTLR{\_}MPS project also uses ANTLRv4 grammar notation~\cite{ANTLR4} and tries to import grammars inside MPS.
However, it does not try to generate the projectional editor at all (i.e., it does not deal with this problem), probably because it is in an early stage of development.

The way this import process of ANTLR{\_}MPS works is quite different from what we have seen in the PE4MPS project, and it is quite complex/complicated to use.
It works as follows. We give only a brief high-level overview and omit low-level technical details.
The author created a whole new ANTLRv4 MPS language, which is an MPS port of the grammar notations' syntax.
To import a language, the user utilizes this MPS ANTLR language.
The textual grammar is imported automatically into MPS taking the form of the MPS's ported grammar language (so that the textual grammar is converted into MPS nodes, that means an AST node is created for each grammar rule).
No child-parent relationships in the structure are generated by the tool --- all must be created manually by the user.
There are no editor nor TextGen aspects created neither.

\paragraph{mps-metabnf.}
\todo{Popis na zaklade emailove diskuze (Nizozemci).}

\section{Creating Language Definition in MPS}

From all the available notations in which grammars can be written, we decided to use the ANTLRv4 grammar notation~\cite{ANTLR4reference} as the input format (in which the input programming languages have to be defined) for the following reasons.
\begin{itemize}
	\item It is a very widely used grammar notation.
	\item All popular and widely used programming languages have their syntax captured (represented) using this notation --- see the repository~\cite{ref:ANTLR4grammars}.
	\item There is a whole framework/toolset for generating parsers and editing grammars built around it~\cite{ref:ANTLR4}.
	\item It supports Java (it is written in Java), which is the implementation language of MPS and also the default BaseLanguage.
	\item It uses (is based on) an EBNF (Extended Backus-Naur Form) \footnote{https://en.wikipedia.org/wiki/Extended{\_}Backus\%E2\%80\%93Naur{\_}Form} form that it useful because it meets/respects the object oriented design that is used when describing MPS aspects (definitions of MPS languages).
	\item It has its own syntax captured in its own syntax\footnote{https://github.com/antlr/grammars-v4/tree/master/antlr4} (there is an ANTLRv4 grammar for the ANTLRv4 notation). This is a very important feature as it allows us to create a parser for ANTLRv4 grammars so that we can parse grammars of other languages.
\end{itemize}

We use ANTLRv4 grammars as inputs (and work with them).
	https://github.com/antlr/antlr4/blob/master/doc/index.md
	https://github.com/antlr/grammars-v4

Unlike some of the related projects, we did not add any new features into the grammar notation.
We tried to import grammars as they are.

The generation/import process performed by INGRID consists of three phases (import of structure, usable editor, text generation).
We will talk about all phases of the generation/import process, describing encountered challenges/obstacles using an example language.

To fulfill our goal of maximal possible automation, we import/generate the full language structure and then do a lot of mundane and time-consuming work connected to creating projectional editors.

Now we describe/introduce the main steps of our approach (that INGRID tool performs): parsing the grammar, creating the structure, editor, and textgen.
We are able to create/generate the structure aspect and then very basic editor and some default/basic textgen.
In particular, we did not port the ANTLR notation as an MPS language.
In the following subsections, we provide some details about our approach to parsing the input grammar and creating the main aspects of each element (each type of AST node) of an MPS language (structure, editor, and textgen).

During our work on this project, we also discovered/encountered challenges/issues related to input grammars --- intermediate rules in the syntax hierarchy that allow humans to understand the grammar better (structure aspect/component), whitespace (editor aspect).
There is a challenge/issue that we have discovered during our work on this project (in the process of our own implementation), and that is related to construction of the structure aspect.
Grammars are written by humans and therefore contain some structures (intermediary layers) inside of them that help a human brain to understand it better.
However, these structures add unnecessary levels to its syntax hierarchy which lead to problems when using this language inside MPS.
We describe/discuss these challenges in more detail (also using examples) and our solutions to the challenges inside the respective following subsections.

Result of the construction of a MPS language from the input grammar (import) has to be polished by the human user.
For example, the aspects not supported by INGRID will be used for adjusting the imported language and improving the usability of the language.

\todo{Briefly discuss how our overall approach differs from similar projects (PE4MPS, etc).}
There are more ways how to approach the problem of a grammar import.

Our overall approach differs from the PE4MPS project (Section X.Y) especially in the level of automation.
While the main goal of (our) the INGRID approach is to automatize the construction of MPS languages from ANTLR grammars as much as possible, usage of PE does not really spare the user any work because it only shifts the manual work from creating projectional editor to writing PE grammars.
What the user would do inside the MPS UI using the projectional editor is in the case of PE4MPS done using a text editor when editing PE grammars.
MPS UI is specially designed/made/tailored for this purpose, guides the user through this process, and supports aids such as auto-completion. In particular, MPS UI allows the user to perform only changes that result in valid output.
Editing the grammar stored in a text file using a standard editor is an inferior approach with respect to using MPS (projectional editor), because a text editor is not as user-friendly, and, more importantly, it is an error prone environment.

Main drawback of ANTLR{\_}MPS, and main difference from our approach, is very limited automation: it does not create the parent-child relationships between AST nodes in the structure, and does not generate any editor and textgen aspects (for any node) at all.
In addition, there is no attention paid to the problem of generating projectional editor and text generation is neglected too.
We blame all this on the project being in its early stage of development.

In the rest of this section, first we define the language that we will use throughout the paper as the running example.
Then we will talk about parsing grammar files \todo{Tohle je mozna zbytecne}.
Lastly, we will describe our approach to tackling generation of language's aspects (structure, editor, and TextGen).
At some places, we discuss important possible alternatives (as there are several possible approaches).

\subsection{Running Example}

We illustrate the main steps on the example of a simplified XML language, which is defined below (in Figure XY).

\todo{Pouzit jazyk SimpleXML na running exaple.}

We decided to use SimpleXML because it is small enough to fit here, but still can be used to illustrate the main challenges and our solution

Throughout the whole paper, we will be using one example language to illustrate main challenges and our solution.
The language is a simplified version of XML. We call it \emph{SimpleXML}.
We created the grammar of the language by taking the official XML ANTLRv4 grammar\footnote{https://github.com/antlr/grammars-v4/tree/master/xml} and stripped it down to eliminate/avoid some not so interesting (advanced) features such as XML entities.
Below in Figure X, you can find the grammar of the SimpleXML language in the ANTLRv4 notation.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\textbf{grammar} \textit{SimpleXML};
	\antlrparserrule{document}     :   \antlrparserrule{prolog}? \antlrparserrule{comment}? \antlrparserrule{element} ;
	\antlrparserrule{prolog}       :   \antlrliteral{<?xml } \antlrparserrule{attribute}* \antlrliteral{?>} ;
	\antlrparserrule{comment}      :   \antlrliteral{<!--} \antlrlexerrule{TEXT} \antlrliteral{-->} ;
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
	             |   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{/>}
	             ;
	\antlrparserrule{attribute}    :   \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT} \antlrliteral{"} ;
	\antlrparserrule{content}      :   \antlrlexerrule{TEXT}
	             |   \antlrparserrule{element}
	             |   \antlrparserrule{comment}
	             |   \antlrlexerrule{CDATA}
	             ;
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
	             |   \antlrliteral{-} | \antlrliteral{_} | \antlrliteral{.}
	             |   \antlrlexerrule{DIGIT}
	             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
	\antlrlexerrule{TEXT}         :   \antlrregex{~[<"]*} ;
	\antlrlexerrule{CDATA}        :   \antlrliteral{<![CDATA[} \antlrregex{.*?} \antlrliteral{]]>} ;
\end{antlr}
\caption{SimpleXML grammar}
\end{figure*}

We have colored the grammar so it is easier to find the way around it (to make navigation easier).
The grammar contains following basic elements:

\begin{itemize}
	\item \textbf{ANTLRv4 keywords} -- In black, they are required by the specification.
	\item \antlrparserrule{\textbf{Parser rules}} -- Stated in blue, they are the rules describing language's structure.
		The alternatives (that the element on the left side may break into) are separated by the pipe ($|$) character.
	\item \antlrlexerrule{\textbf{Lexer rules}} -- In yellow, they are rules describing terminal symbols that the parser is matching against the input.
		Terminals are encoded as string values and regular expressions.
		The \textbf{fragment} keyword states that the lexer sub-rule is just a helper rule that brings more clarity into the notation but is not visible in the parser output.
	\item \antlrliteralnoap{\textbf{String values (literals)}} -- In red, always inside of a pair of apostrophes, they are similar to lexer rules.
		They describe a terminal string, not using regular expressions but by an exact string match.
	\item \antlrregex{\textbf{Regular expressions}} -- Colored in green, they describe a string token to be matched using a regular expression with the special ANTLRv4 regex notation\footnote{https://github.com/antlr/antlr4/blob/master/doc/lexer-rules.md}.
	\item \textbf{Rule operators (?+*)} - Left in black, following any element, they quantify the number of occurrences the prepending element can appear in.
		These operators can also be prepended by an asterisk (*), meaning they are non-greedy.
\end{itemize}

\subsection{Parsing (the) Grammar}

\todo{Define some name for the AST representing the grammar, and use it consistently throughout the whole paper. Maybe distinguish between "grammar AST" and "MPS structure AST".}
The first step (stage) of creating the MPS language (of the import process) is parsing the grammar file (written using the ANTLRv4 notation) so that we can get an AST tree representing the grammar.

The abstract syntax tree that comes out of the automatically generated (ANTLR) parser is a little bit complicated and full of information not relevant for our purposes (to us).
That is why we translate the complex ANTLR AST into our own simplified tree made out of our own objects.
We discarded as much irrelevant information as possible and only kept information (data) vital/necessary for construction of MPS languages.

Furthermore, we process the grammar in two iterations.
In the first iteration/pass, we build a tree holding names of other referenced rules in the form of strings.
In the second iteration, we walk this tree and resolve all these references to real pointers to other rule objects.
We end up with a clear and simpler representation that is easy to process in the next steps of the import process.

We also had to post-process (simplify) the representation of lexer rules (tokens).
As a result of the initial parsing, ANTLR parser produces some kind of a tree structure that is representing the lexer rule (meaning/note that lexer rules can also be built from alternatives just like parser rules).
However, lexer rules are, in fact (basically), just regular expressions used for parsing input source code into tokens.
We took each/every such tree that represents the lexer rule and flattened it into the equivalent regular expression.

For illustration, consider the lexer rule \antlrlexerrule{Name} from our/the SimpleXML language (Figure XY), together with all its sub-rules, shown in Figure~\ref{fig:lexerrule}.

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrlexerrule{Name}         :   \antlrlexerrule{NameStartChar} \antlrlexerrule{NameChar}* ;
	\textbf{fragment}
	\antlrlexerrule{DIGIT}        :   \antlrregex{[0-9]} ;
	\textbf{fragment}
	\antlrlexerrule{NameChar}     :   \antlrlexerrule{NameStartChar}
             |   \antlrliteral{-} | \antlrliteral{{\_}} | \antlrliteral{.}
             |   \antlrlexerrule{DIGIT}
             ;
	\textbf{fragment}
	\antlrlexerrule{NameStartChar}:   \antlrregex{[:a-zA-Z]} ;
\end{antlr}
\caption{Lexer rule from SimpleXML}
\label{fig:lexerrule}
\end{figure*}

Intuitively, we can create a regular expression by gluing elements of each alternative together and then join these alternatives with a classic regex OR ($|$), which is exactly its function in the ANTLRv4 grammar notation.
We have captures this idea in the form of the recursive algorithm in Figure~\ref{fig:algflatten}, which covers the whole process.

% TODO prepsat jako standardni pseudocode (nejspis pouzit jiny latex environment)
\begin{figure*}[ht]
\centering
\begin{antlr}
	Flatten(\textit{R}):
	1) Define \textit{T} as a list of empty string
	2) For each alternative \textit{A} of the rule \textit{R}:
	3)     Define \textit{R} = \antlrap\antlrap
	4)     For each element \textit{E} of \textit{A}:
	5)         If \textit{E} is not yet flattened:
	6)             Flatten(\textit{E})
	7)         \textit{R}.append(\textit{E})
	8)         \textit{R}.append(\textit{E}.operator)
	9)     \textit{T}.add(\textit{R})
	10) Build a string S out of elements t$_{\text{1}}$, t$_{\text{2}}$, ... t$_{\text{n}}$ of \textit{T}:
	11)     (t$_{\text{1}}$)|(t$_{\text{2}}$)|...|(t$_{\text{n}}$)
	12) Return \textit{S}
\end{antlr}
\caption{Flatten algorithm}
\label{fig:algflatten}
\end{figure*}

The output of \texttt{Flatten(Name)} (output of the Flatten algorithm applied to the \antlrlexerrule{Name} rule) is this regular expression (in this form):

\begin{center}
	\texttt{([:a-zA-Z])(([:a-zA-Z])|(-)|({\_})|(.)|([0-9]))*}
\end{center}

\subsection{Structure}

Now, as the next step, we need to translate/transform the AST (the types of AST nodes) derived from the input grammar into elements of the structure aspect of the MPS language definition (into types of AST nodes), and linking the nodes appropriately.

The goal of this second step is to translate (transfer) the syntax (structure) of the input language that is captured by the grammar (ANTLR), into the MPS structure aspect.
The structure of the MPS language is somewhat similar to the grammar.
This is (the part) where INGRID spares the user from many hours of tedious and sometimes quite challenging work.
The process of transferring grammar rules into the structure of a MPS language presents/involves some interesting challenges.
The input grammar (EBNF, ANTLR) serves a little bit different purpose than MPS language, and the way grammar rules are usually written down causes some problems that we had to overcome in order to recreate the language inside MPS.
Note that the challenges and problems are not specific to MPS --- they are common for everyone tackling a similar problem.

We are going to describe two different approaches that we designed --- the straightforward approach (Section~\ref{sect:straightforward_approach}) and the shortcut approach (Section~\ref{sect:shortcut_approach}).
We will describe two possible approaches that we developed in Section X and Section Y, respectively.
\todo{Rict neco jako ze ten druhy, slozitejsi, approach vlastne rozsiruje ten prvni, jednodussi.}

However, before focusing on the two approaches, we must describe the basic principles of translation from the AST representing the grammar to the MPS structure, which are common to both approaches.
The structure is created according to object-oriented principles.
Each node (element) is represented by an object (and can be treated as such) that may have a parent node, child nodes, and properties of any data type (similar to fields), it may implement any number of interfaces, and it may also hold references to other nodes.
For every rule of the grammar (for every node in the grammar AST) that breaks into more/multiple alternatives, we must create a node in the MPS structure AST for each alternative of the grammar (every child node).
Names of the nodes for alternatives are composed from the name of the rule and a number indicating the order of appearance in the grammar.
For example, considering the fragment of the grammar of the SimpleXML language in Figure~\ref{fig:xmlgrammarelement}

\begin{figure*}[ht]
\centering
\begin{antlr}
	\antlrparserrule{element}      :   \antlrliteral{<} \antlrlexerrule{Name} \antlrparserrule{attribute}* \antlrliteral{>} \antlrparserrule{content}* \antlrliteral{</} \antlrlexerrule{Name} \antlrliteral{>}
             \textcolor{gray}{|   \antlrap<\antlrap Name attribute* \antlrap/>\antlrap}
             \textcolor{gray}{;}
\end{antlr}
\caption{Fragment of the SimpleXML grammar}
\label{fig:xmlgrammarelement}
\end{figure*}

Results of this step for the first alternative, which represents the represents the full XML tag with content, can be seen in Figure~\ref{fig:element_node_common}.

\begin{figure}[ht]
	\centering
	\includegraphics[scale=0.75]{./images/element_node_common.png}
	\caption{Resulting Element{\_}1 node}
	\label{fig:element_node_common}
\end{figure}

String literals (e.g., keywords such as \textbf{for}) are omitted/skipped because they will appear only in the corresponding projectional editor.
The node named \mpsconcept{Element{\_}1} contains two properties, one for each reference to the lexer rule \antlrlexerrule{Name} inside the first alternative of the \antlrparserrule{element} rule,
Value of the property will be restricted using the regular expression representing the \antlrlexerrule{Name} rule.
We explain later how the references to parser rules (\antlrparserrule{attribute} and \antlrparserrule{content}) as this is where the approaches differ.
The main/key difference between these two lies in the way we create children fields for parser rules and in the way we are going to link them together using interface concepts.

\subsubsection{The Straightforward Approach}
\label{sect:straightforward_approach}

>>>>

% TODO pridat nekoho dalsiho z JetBrains (Alex Shatalin?)
% TODO upravit affiliation pro JetBrains -> doplnit "s.r.o" nebo radsi "Inc."
% TODO muzeme taky napsat jejich emaily pokud budou lidi z JetBrains souhlasit
\authorinfo{V\'aclav Pech}
	{JetBrains}

@@

\input{chapters/introduction}
\input{chapters/background}
\input{chapters/language_definition}
\input{chapters/evaluation}
\input{chapters/conclusion}
